Concurrent Programming with Harmony

### 2020-07-28 book/paper.tex
Finished pass over chapter 7

Harmony is not an object-oriented language like Python is.  In Python
you can pass a reference to an object to a method, and that method
can then update the object.  In Harmony, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this, as mentioned in \autoref{ch:harmonymachine},
each shared variable has an \emph{address}.
\index{address}%
If $x$ is a shared variable, then the expression ?$x$ is the address of $x$.
If a variable contains an address, we call that variable a \emph{pointer}.
\index{pointer}%
If $p$ is a pointer to a shared variable, then the
expression \texttt{!{}}$p$ is the value of the shared variable.
In particular, \texttt{!{}?}$x$ == $x$.
This is similar to how C pointers work (\texttt{*\string&}$x$ == $x$).

\autoref{fig:petersonmethods} again shows Peterson's algorithm,
but this time with methods defined to enter and exit the critical
section.
The name \textit{mutex} is often used to denote a variable or value
that is used for mutual exclusion.
\texttt{P\_mutex} is a method that returns a ``mutex,'' which, in this
case, is a dictionary that contains Peterson's Algorithm's state:
a turn variable and two flags.
Both methods \texttt{P\_enter} and \texttt{P\_exit} take two arguments:
a pointer to a mutex and the process identifier (0 or 1).
\texttt{(!{}}\textit{pm}\texttt{.turn} is the value of the \texttt{.turn} key
in the dictionary that \textit{pm} points to. 
Note that the parentheses in this expression
are needed, as \texttt{!{}}\textit{pm}\texttt{.turn} would wrongly evaluate to
\texttt{!{}}(\textit{pm}\texttt{.turn}).

@@

$\mathcal{I}_1$ is an inductive invariant.
To see that invariant $\mathcal{I}_1$ is maintained, note that
\texttt{!{}}$p$ == \texttt{True} upon entry of \texttt{tas}
(because of the condition on the \textbf{while} loop that the
\texttt{tas} method is invoked in).
So there are two cases:
\begin{enumerate}
\item \texttt{!{}}$s$ is \texttt{False} upon entry to \texttt{tas}.
Then upon exit \texttt{!{}}$p$ == \texttt{False} and \texttt{!{}}$s$ == \texttt{True},
maintaining the invariant.
\item \texttt{!{}}$s$ is \texttt{True} upon entry to \texttt{tas}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section.
Invariant $\mathcal{I}_2(i)$ is obvious as (i) process $i$ only proceeds to the critical
section if \textit{private}[$i$] == \texttt{False}, and (ii) no other process modifies
\textit{private}[$i$].

@@

Because of the network's properties, we can model each
channel as a variable that either contains a network message or nothing
(we use \texttt{()} in the model).  Let \textit{s\_chan} be the channel
from Alice to Bob and \textit{r\_chan} the channel from Bob to Alice.
\texttt{net\_send}(\textit{pchan}, $m$, \textit{reliable}) models sending
a message $m$ to
\texttt{!{}}\textit{pchan}, where \textit{pchan} is either ?\textit{s\_chan}
or ?\textit{r\_chan}.
The method places either $m$ (to model a successful send)
or \texttt{()} (to model loss) in \texttt{!{}}\textit{pchan}.
The use of the \textit{reliable} flag will be explained later.
\texttt{net\_recv}(\textit{pchan}) models checking \texttt{!{}pchan} for
the next message.

@@

As a first example, we will revisit the producer/consumer problem.
The code in \autoref{fig:hw} is based on code developed by Herlihy and
Wing~\cite{HW87}.
The code is a ``proof of existence'' for non-blocking synchronization; it
is not necessarily practical.
There are two variables.  \textit{items} is an unbounded array with each
entry initialized to \texttt{()}.  \textit{back} is an index into the
array and points to the next slot where a new value is inserted.
The code uses two interlock instructions:
\begin{itemize}
\item \texttt{inc}($p$): atomically increments \texttt{!{}}$p$ and returns
the old value;
\item \texttt{swap}($p$): sets \texttt{!{}}$p$ to \texttt{()} and returns
the old value.
\end{itemize}

@@

{\small
\begin{tabular}{|l|l|}
\hline
\texttt{$e$ == $e$}, \texttt{$e$ != $e$} & two Harmony values of the same type \\
\hline
\texttt{$e$ < $e$, $e$ <= $e$, $e$ > $e$, $e$ >= $e$} & any two Harmony values \\
\hline
\texttt{$e$ and $e$ and ...} & two or more booleans \\
\hline
\texttt{$e$ or $e$ or ...} & two or more booleans \\
\hline
\texttt{not $e$} & a boolean \\
\hline
\texttt{-$e$, \string~$e$} & an integer \\
\hline
\texttt{$e$ + $e$ + ...} & two or more integers, tuples, or lists \\
\hline
\texttt{$e$ - $e$} & two integers or sets \\
\hline
\texttt{$e$ \& $e$ \& ...} & two or more integers or sets \\
\hline
\texttt{$e$ | $e$ | ...} & two or more integers or sets \\
\hline
\texttt{$e$ \string^ $e$ \string^ ...} & two or more integers or sets \\
\hline
\texttt{$e$ * $e$} & two integers or an integer and a list \\
\hline
\texttt{$e$ / $e$, $e$ // $e$} & two integers (integer division) \\
\hline
\texttt{$e$ \% $e$, $e$ mod $e$} & two integers (division remainder) \\
\hline
\texttt{$e$ ** $e$, $e << e$, $e >> e$} & two integers \\
\hline
\texttt{$\{ e..e \}$} & two integers \\
\hline
\texttt{$e$ [not] in $e$} & first is any Harmony value, second is set \\
\hline
\texttt{$e$ if $e$ else $e$} & middle $e$ must be a boolean \\
\hline
\texttt{!{}$e$} & an address \\
\hline
\texttt{choose $e$} & a set \\
\hline
\texttt{min $e$, max $e$, len $e$} & a set or a dict \\
\hline
\texttt{keys $e$} & a dictionary (includes tuple and list) \\
\hline
\texttt{bagsize $e$} & a bag \\
\hline
\texttt{atLabel $e$} & atom (corresponding to a label) \\
\hline
\texttt{nametag $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{processes $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{setintlevel} \textit{e} & \textit{e} is a boolean \\
\hline
\texttt{stop} \textit{lv} & \textit{lv} is an lvalue \\
\hline
\texttt{?}\textit{lv} & \textit{lv} is an lvalue \\
\hline
\texttt{lambda $a$: $e$ end} & $a$ is a bounded variable, $e$ an expression\\
\hline
\end{tabular}
}
\vspace{1em}

@@

An \emph{lvalue}
\index{lvalue}%
(short for left hand value of an assignment statement)
is something that can be assigned.  This can be a
shared variable, a process variable, or a dereferenced pointer
variable.  It can also be indexed.  Examples include
\textit{var}, \textit{var}[$e$], and \texttt{!{}}\textit{p}.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{bagEmpty()} & returns an empty bag\\
\hline
\texttt{bagFromSet($s$)} & create a bag from set $s$\\
\hline
\texttt{bagCount($b$, $e$)} & count how many times $e$ occurs in bag $b$\\
\hline
\texttt{bagChoose($b$)} & like \texttt{choose(s)}, but applied to a bag\\
\hline
\texttt{bagAdd}(\textit{pb}, $e$) & add one copy of $e$ to bag \texttt{!{}}\textit{pb}\\
\hline
\texttt{bagRemove}(\textit{pb}, $e$) & remove one copy of $e$ from bag \texttt{!{}}\textit{pb}\\
\hline
\end{tabular}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{tas}(\textit{lk}) & test-and-set on \texttt{!{}}\textit{lk} \\
\hline
\texttt{Lock}() & return a lock object \\
\hline
\texttt{lock}(\textit{lk}) & acquire \texttt{!{}}\textit{lk} \\
\hline
\texttt{unlock}(\textit{lk}) & release \texttt{!{}}\textit{lk} \\
\hline
\texttt{Condition}(\textit{lk}) & return a condition variable on \texttt{!{}}\textit{lk} \\
\hline
\texttt{wait}($c$) & wait on condition variable \texttt{!{}}$c$ \\
\hline
\texttt{notify}($c$) & notify a process waiting on condition variable \texttt{!{}}$c$ \\
\hline
\texttt{notifyAll}($c$) & notify all processes waiting on condition variable \texttt{!{}}$c$ \\
\hline
\texttt{Semaphore}(\textit{cnt}) & return a semaphore object initialized to \textit{cnt} \\
\hline
\texttt{P}(\textit{sema}) & procure \texttt{!{}}\textit{sema}  \\
\hline
\texttt{V}(\textit{sema}) & vacate \texttt{!{}}\textit{sema}  \\
\hline
\texttt{Queue}() & return a synchronized queue object \\
\hline
\texttt{dequeue}($q$) & return next element of $q$, blocking if empty \\
\hline
\texttt{enqueue}($q$, \textit{item}) & add \textit{item} to $a$ \\
\hline
\end{tabular}

>>>>

Harmony is not an object-oriented language like Python is.  In Python
you can pass a reference to an object to a method, and that method
can then update the object.  In Harmony, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this, as mentioned in \autoref{ch:harmonymachine},
each shared variable has an \emph{address}.
\index{address}%
If $x$ is a shared variable, then the expression ?$x$ is the address of $x$.
If a variable contains an address, we call that variable a \emph{pointer}.
\index{pointer}%
If $p$ is a pointer to a shared variable, then the
expression \texttt{!}$p$ is the value of the shared variable.
In particular, \texttt{!?}$x$ == $x$.
This is similar to how C pointers work (\texttt{*\string&}$x$ == $x$).

\autoref{fig:petersonmethods} again shows Peterson's algorithm,
but this time with methods defined to enter and exit the critical
section.
The name \textit{mutex} is often used to denote a variable or value
that is used for mutual exclusion.
\texttt{P\_mutex} is a method that returns a ``mutex,'' which, in this
case, is a dictionary that contains Peterson's Algorithm's state:
a turn variable and two flags.
Both methods \texttt{P\_enter} and \texttt{P\_exit} take two arguments:
a pointer to a mutex and the process identifier (0 or 1).
(\texttt{!}\textit{pm})\texttt{.turn} is the value of the \texttt{.turn} key
in the dictionary that \textit{pm} points to. 
Note that the parentheses in this expression
are needed, as \texttt{!}\textit{pm}\texttt{.turn} would wrongly evaluate
\texttt{!}(\textit{pm}\texttt{.turn}).

@@

$\mathcal{I}_1$ is an inductive invariant.
To see that invariant $\mathcal{I}_1$ is maintained, note that
\texttt{!}$p$ == \texttt{True} upon entry of \texttt{tas}
(because of the condition on the \textbf{while} loop that the
\texttt{tas} method is invoked in).
So there are two cases:
\begin{enumerate}
\item \texttt{!}$s$ is \texttt{False} upon entry to \texttt{tas}.
Then upon exit \texttt{!}$p$ == \texttt{False} and \texttt{!}$s$ == \texttt{True},
maintaining the invariant.
\item \texttt{!}$s$ is \texttt{True} upon entry to \texttt{tas}.
Then upon exit nothing has changed, maintaining the invariant.
\end{enumerate}
Invariant $\mathcal{I}_1$ is also easy to verify for exiting the critical section.
Invariant $\mathcal{I}_2(i)$ is obvious as (i) process $i$ only proceeds to the critical
section if \textit{private}[$i$] == \texttt{False}, and (ii) no other process modifies
\textit{private}[$i$].

@@

Because of the network's properties, we can model each
channel as a variable that either contains a network message or nothing
(we use \texttt{()} in the model).  Let \textit{s\_chan} be the channel
from Alice to Bob and \textit{r\_chan} the channel from Bob to Alice.
\texttt{net\_send}(\textit{pchan}, $m$, \textit{reliable}) models sending
a message $m$ to
\texttt{!}\textit{pchan}, where \textit{pchan} is either ?\textit{s\_chan}
or ?\textit{r\_chan}.
The method places either $m$ (to model a successful send)
or \texttt{()} (to model loss) in \texttt{!}\textit{pchan}.
The use of the \textit{reliable} flag will be explained later.
\texttt{net\_recv}(\textit{pchan}) models checking \texttt{!pchan} for
the next message.

@@

As a first example, we will revisit the producer/consumer problem.
The code in \autoref{fig:hw} is based on code developed by Herlihy and
Wing~\cite{HW87}.
The code is a ``proof of existence'' for non-blocking synchronization; it
is not necessarily practical.
There are two variables.  \textit{items} is an unbounded array with each
entry initialized to \texttt{()}.  \textit{back} is an index into the
array and points to the next slot where a new value is inserted.
The code uses two interlock instructions:
\begin{itemize}
\item \texttt{inc}($p$): atomically increments \texttt{!}$p$ and returns
the old value;
\item \texttt{swap}($p$): sets \texttt{!}$p$ to \texttt{()} and returns
the old value.
\end{itemize}

@@

{\small
\begin{tabular}{|l|l|}
\hline
\texttt{$e$ == $e$}, \texttt{$e$ != $e$} & two Harmony values of the same type \\
\hline
\texttt{$e$ < $e$, $e$ <= $e$, $e$ > $e$, $e$ >= $e$} & any two Harmony values \\
\hline
\texttt{$e$ and $e$ and ...} & two or more booleans \\
\hline
\texttt{$e$ or $e$ or ...} & two or more booleans \\
\hline
\texttt{not $e$} & a boolean \\
\hline
\texttt{-$e$, \string~$e$} & an integer \\
\hline
\texttt{$e$ + $e$ + ...} & two or more integers, tuples, or lists \\
\hline
\texttt{$e$ - $e$} & two integers or sets \\
\hline
\texttt{$e$ \& $e$ \& ...} & two or more integers or sets \\
\hline
\texttt{$e$ | $e$ | ...} & two or more integers or sets \\
\hline
\texttt{$e$ \string^ $e$ \string^ ...} & two or more integers or sets \\
\hline
\texttt{$e$ * $e$} & two integers or an integer and a list \\
\hline
\texttt{$e$ / $e$, $e$ // $e$} & two integers (integer division) \\
\hline
\texttt{$e$ \% $e$, $e$ mod $e$} & two integers (division remainder) \\
\hline
\texttt{$e$ ** $e$, $e << e$, $e >> e$} & two integers \\
\hline
\texttt{$\{ e..e \}$} & two integers \\
\hline
\texttt{$e$ [not] in $e$} & first is any Harmony value, second is set \\
\hline
\texttt{$e$ if $e$ else $e$} & middle $e$ must be a boolean \\
\hline
\texttt{!$e$} & an address \\
\hline
\texttt{choose $e$} & a set \\
\hline
\texttt{min $e$, max $e$, len $e$} & a set or a dict \\
\hline
\texttt{keys $e$} & a dictionary (includes tuple and list) \\
\hline
\texttt{bagsize $e$} & a bag \\
\hline
\texttt{atLabel $e$} & atom (corresponding to a label) \\
\hline
\texttt{nametag $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{processes $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{setintlevel} \textit{e} & \textit{e} is a boolean \\
\hline
\texttt{stop} \textit{lv} & \textit{lv} is an lvalue \\
\hline
\texttt{?}\textit{lv} & \textit{lv} is an lvalue \\
\hline
\texttt{lambda $a$: $e$ end} & $a$ is a bounded variable, $e$ an expression\\
\hline
\end{tabular}
}
\vspace{1em}

@@

An \emph{lvalue}
\index{lvalue}%
(short for left hand value of an assignment statement)
is something that can be assigned.  This can be a
shared variable, a process variable, or a dereferenced pointer
variable.  It can also be indexed.  Examples include
\textit{var}, \textit{var}[$e$], and \texttt{!}\textit{p}.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{bagEmpty()} & returns an empty bag\\
\hline
\texttt{bagFromSet($s$)} & create a bag from set $s$\\
\hline
\texttt{bagCount($b$, $e$)} & count how many times $e$ occurs in bag $b$\\
\hline
\texttt{bagChoose($b$)} & like \texttt{choose(s)}, but applied to a bag\\
\hline
\texttt{bagAdd}(\textit{pb}, $e$) & add one copy of $e$ to bag \texttt{!}\textit{pb}\\
\hline
\texttt{bagRemove}(\textit{pb}, $e$) & remove one copy of $e$ from bag \texttt{!}\textit{pb}\\
\hline
\end{tabular}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{tas}(\textit{lk}) & test-and-set on \texttt{!}\textit{lk} \\
\hline
\texttt{Lock}() & return a lock object \\
\hline
\texttt{lock}(\textit{lk}) & acquire \texttt{!}\textit{lk} \\
\hline
\texttt{unlock}(\textit{lk}) & release \texttt{!}\textit{lk} \\
\hline
\texttt{Condition}(\textit{lk}) & return a condition variable on \texttt{!}\textit{lk} \\
\hline
\texttt{wait}($c$) & wait on condition variable \texttt{!}$c$ \\
\hline
\texttt{notify}($c$) & notify a process waiting on condition variable \texttt{!}$c$ \\
\hline
\texttt{notifyAll}($c$) & notify all processes waiting on condition variable \texttt{!}$c$ \\
\hline
\texttt{Semaphore}(\textit{cnt}) & return a semaphore object initialized to \textit{cnt} \\
\hline
\texttt{P}(\textit{sema}) & procure \texttt{!}\textit{sema}  \\
\hline
\texttt{V}(\textit{sema}) & vacate \texttt{!}\textit{sema}  \\
\hline
\texttt{Queue}() & return a synchronized queue object \\
\hline
\texttt{dequeue}($q$) & return next element of $q$, blocking if empty \\
\hline
\texttt{enqueue}($q$, \textit{item}) & add \textit{item} to $a$ \\
\hline
\end{tabular}
